Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-12 treestats_paper.tex

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \log n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove after the initial building of the tree
may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$.
Therefore, the overall running time
of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
\[
    O\left(n + \rho \log(n)^2 \right)
\]
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

\begin{figure}
    % Run the script illustrate_algorithm.py to generate these pdfs and
    % the latex tables below.
    \begin{center}
    \begin{tabular}{ccc}
    \includegraphics[width=4cm]{figures/tree_0_init.pdf} & &
    \includegraphics[width=4cm]{figures/tree_0_out_0.pdf}
    \\

@@

    \caption{
    Illustration of the internal state for algorithm B for $\iw = \bone_S$ and
    $f(x) = x(5 - x) / 20$ as in Example~\ref{ex:site_diversity}. (A)
    Immediately before we remove the edge joining 6 to 7 and (B) immediately after.
    See the text for descriptions of the arrays encoding the state.
    \jk{TODO: the value of $s$ would be helpful here too.}
    \label{fig:algorithm_example}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Speed}

\krt{need to sort out the big-O issue here}
We used coalescent simulations to compare the performance of calculating Tajima's \citeyearpar{Tajima1989-de} $D$
statistic from tree sequences to calculating it from integer matrices containing genotypes at all variable positions.
\autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} processes variant data
substantially faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids --
three times faster for one thousand haplotypes,
growing to nearly three orders of magnitude faster for one million haplotypes.
The expected number of mutations for each replicate is $10^4 \times \sum_{i=1}^{n-1}\frac{1}{i}$, and thus ranges from
$\approx 32,000$ to $\approx 138,000$ \citep{Watterson1975-ej}.
\autoref{fig:stats_performance} only considers the time required to calculate the statistic,
and does not include the time required to build the genotype matrix, nor the memory required to store it.
For the largest samples size of $n=10^6$,
the matrix is approximately 200 gigabytes, and thus not practical on many systems.
The largest tree sequence simulated required less than one gigabyte of memory.

For larger sample sizes, there is an uptick in the time required per variant
by \texttt{tskit} shown in \autoref{fig:stats_performance}.
This occurs because the $O(n)$ complexity of building the first tree
begins to dominate run time for large $n$,
a cost that is amortized in longer genomes where more recombination leads to more trees.

\begin{figure}
    \centering
    \includegraphics{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
    \caption{Comparison of time required to compute site statistics
        between matrix and tree based methods. For each sample size, a single replicate
        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$,
        where $\mu$ and $\rho$ are the total mutation and recombination rate across the simulated region
        per generation, respectively.
        For each replicate, Tajima's $D$
        \citep{Tajima1989-de} statistic was calculated using \tskit, \texttt{libsequence}
        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the
        algorithm used by \texttt{tskit}. 
        \label{fig:stats_performance}
    }
\end{figure}

>>>>

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \log n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove after the initial building of the tree
may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$.
Therefore, the overall running time
of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
\begin{equation}\label{eqn:alg_complexity}
    O\left(n + \rho \log(n)^2 \right)
\end{equation}
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

\begin{figure}
\footnotesize
    % Run the script illustrate_algorithm.py to generate these pdfs and
    % the latex tables below.
    \begin{center}
    \begin{tabular}{ccc}
    \includegraphics[width=4cm]{figures/tree_0_init.pdf} & &
    \includegraphics[width=4cm]{figures/tree_0_out_0.pdf}
    \\
    $s = 3.1$ & & $s = 2.9$\\

@@

    \caption{
    Illustration of the internal state for algorithm B for $\iw = \bone_S$ and
    $f(x) = x(5 - x) / 20$ as in Example~\ref{ex:site_diversity}. (A)
    Immediately before we remove the edge joining 6 to 7 and (B) immediately after.
    See the text for descriptions of the arrays encoding the state. Also shown
    is the value of the running sum, $s = \sum_u \beta_u F_u$.
    \label{fig:algorithm_example}}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Efficiency}

We used coalescent simulations to compare the performance of calculating Tajima's \citeyearpar{Tajima1989-de} $D$
statistic from tree sequences to calculating it from integer matrices containing genotypes at all variable positions.
\autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} processes variant data
substantially faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids --
three times faster for one thousand haplotypes,
growing to nearly three orders of magnitude faster for one million haplotypes.
The expected number of mutations for each replicate is $10^4 \times \sum_{i=1}^{n-1}\frac{1}{i}$, and thus ranges from
$\approx 32,000$ to $\approx 138,000$ \citep{Watterson1975-ej}.
\autoref{fig:stats_performance} only considers the time required to calculate the statistic,
and does not include the time required to build the genotype matrix, nor the memory required to store it.
For the largest samples size of $n=10^6$,
the matrix is approximately 200 gigabytes, and thus not practical on many systems.
The largest tree sequence simulated required less than one gigabyte of memory.

To determine how well our theoretical model of time complexity predicts the
running time in \tskit{}'s implementation of the site statistics algorithm,
we fitted a model based on Equation~\eqref{eqn:alg_complexity}.
\autoref{fig:stats_performance} shows that theoretical predictions match
the observed running time very well. As the overall complexity is
$O(n + \rho \log(n)^2)$, for sufficiently large $n$ the initial term
(representing the cost of building the first tree) will come to dominate.
In this case, we can see this happening at around $n = 10^5$, where there
is a noticeable uptick in the time required per variant. For
longer genomes (i.e., larger values of $\rho$), this cost is amortized over more trees
and is less apparent.

\begin{figure}
    \centering
    \includegraphics[width=9cm]{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
    \caption{Comparison of time required to compute site statistics
        between matrix and tree based methods. For each sample size, a single replicate
        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$,
        where $\mu$ and $\rho$ are the total mutation and recombination rate across the simulated region
        per generation, respectively.
        For each replicate, Tajima's $D$
        \citep{Tajima1989-de} statistic was calculated using \tskit, \texttt{libsequence}
        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the algorithm used by \texttt{tskit}
        (\autoref{eqn:alg_complexity}).
        \label{fig:stats_performance}
    }
\end{figure}

It should be noted here that this example based on simulated data
represents a best-case scenario in terms of the performance advantages
of \tskit{} over the matrix-based methods. For real data, inferred
tree sequences currently contain substantially more edges than we
would expect based on simple neutral theory~\citep{kelleher2019inferring},
and therefore computing statistics will not be as efficient as
for an equivalent simulation. Tree sequence inference
methods are in their infancy, however, and it is likely that
as they improve the number of edges required to encode data will
be reduced. For data simulated natively in tree sequence format
via packages such as \texttt{msprime}~\citep{kelleher2016efficient},
\texttt{SLiM}~\citep{haller2018tree,haller2019slim} and
\texttt{fwdpy11}~\citep{thornton2014c++}, the advice is straightforward.
Computing statistics using the algorithms in \tskit{}
will always be more efficient than decoding the genotype
matrix, importing it into another package and computing from the matrix.
